# Quality Control

Steps:

1.  How to qc segmentation (simpleSeg/cellSPA?)
2.  How to qc image batch-effect (simpleSeg::normalizeCells)
3.  How to qc patient batch-effect (simpleSeg::normalizeCells)
4.  How to qc batch effects (scMerge)

## CellSPA: How do I determine segmentation quality?

## simpleSeg: Do my images have a batch effect?

In many spatial imaging protocols, there tends to be a degree of variability in the intensity of each image. For example, in one image, the CD3 stain may be too strong, whereas in another image the CD3 staining is particularly weak. This variability is often inevitable and can be hard to correct for during the imaging process. Hence, it is important that we identify when such variance occurs and correct it.

First, let's load in the images we previously segmented out in the last section. The `SpatialDatasets` package conveniently provides the segmented out images for the HNsCC dataset from Ferguson et al., 2022.

```{r, message=FALSE, warning=FALSE}
library(tidySingleCellExperiment)
library(simpleSeg)
# library(scMerge)
library(scater)
library(ggplot2)
```

```{r}
fergusonSPE <- SpatialDatasets::spe_Ferguson_2022()
```

Next, we can check if the marker intensities of each cell require some form of transformation or normalisation. The reason we do this is two-fold:\

1.  The intensities of images are often highly skewed, preventing any meaningful downstream analysis.\
2.  The intensities across different images are often different, meaning that what is considered "positive" can be different across images.

By transforming and normalising the data, we aim to reduce these two effects. Below, we extract the marker intensities from the `counts` assay and take a closer look at rhe CD3 marker, which should be expressed in the majority of T cells.

```{r, fig.width=5, fig.height=5}
# Plot densities of CD3 for each image.
fergusonSPE |> 
  join_features(features = rownames(fergusonSPE), shape = "wide", assay = "counts") |> 
  ggplot(aes(x = CD3, colour = imageID)) + 
  geom_density() + 
  theme(legend.position = "none")
```

Here, we can see that the intensities are very clearly skewed, and it is difficult to distinguish a CD3- cell from a CD3+ cell. Further, we can clearly see some image-level batch effect, where across images, the intensity peaks differ drastically.

Another method of visualising batch effect is using a dimensionality reduction technique and visualising how the images separate out on a 2D plot. If no batch effect is expected, we should see the images largely overlap with each other.

```{r}
# Usually we specify a subset of the original markers which are informative to separating out distinct cell types for the UMAP and clustering.
ct_markers <- c("podoplanin", "CD13", "CD31",
                "panCK", "CD3", "CD4", "CD8a",
                "CD20", "CD68", "CD16", "CD14", 
                "HLADR", "CD66a")

set.seed(51773)
# Perform dimension reduction using UMAP.
fergusonSPE <- scater::runUMAP(
  fergusonSPE,
  subset_row = ct_markers,
  exprs_values = "counts"
)

# Select a subset of images to plot.
someImages <- unique(fergusonSPE$imageID)[c(1, 5, 10, 20, 30, 40)]

# UMAP by imageID.
scater::plotReducedDim(
  fergusonSPE[, fergusonSPE$imageID %in% someImages],
  dimred = "UMAP",
  colour_by = "imageID"
)
```

The UMAP also indicates that some level of batch effect exists in our dataset.

`simpleSeg` provides the `normalizeCells()` function for correcting image-level batch effects. We specify the following parameters -

-   `transformation` is an optional argument which specifies the function to be applied to the data. We do not apply an arcsinh transformation here, as we already apply a square root transform in the `simpleSeg()` function.
-   `method = c("trim99", "mean", PC1")` is an optional argument which specifies the normalisation method/s to be performed. A comprehensive table of methods is provided below.
-   `assayIn = "counts"` is a required argument which specifies the name of the assay that contains our intensity data. In our context, this is called `counts`.

```{r, fig.width=5, fig.height=5}
# Leave out the nuclei markers from our normalisation process. 
useMarkers <- rownames(fergusonSPE)[!rownames(fergusonSPE) %in% c("DNA1", "DNA2", "HH3")]

# Transform and normalise the marker expression of each cell type.
fergusonSPE <- normalizeCells(fergusonSPE,
                        markers = useMarkers,
                        transformation = NULL,
                        method = c("trim99", "mean", "PC1"),
                        assayIn = "counts",
                        cores = nCores
)
```

This modified data is then stored in the `norm` assay by default, but can be changed using the `assayOut` parameter.

#### `method` Parameters

| Method | Description |   |
|------------------|:-------------------------:|:-------------------------:|
| ["mean"]{style="font-family: 'Courier New', monospace;"} | [Divides the marker cellular marker intensities by their mean.]{style="font-family: 'Courier New', monospace;"} |  |
| ["minMax"]{style="font-family: 'Courier New', monospace;"} | [Subtracts the minimum value and scales markers between 0 and 1.]{style="font-family: 'Courier New', monospace;"} |  |
| ["trim99"]{style="font-family: 'Courier New', monospace;"} | [Sets the highest 1% of values to the value of the 99th percentile.\`]{style="font-family: 'Courier New', monospace;"} |  |
| ["PC1"]{style="font-family: 'Courier New', monospace;"} | [Removes the 1st principal component) can be performed with one call of the function, in the order specified by the user.]{style="font-family: 'Courier New', monospace;"} |  |

We can then plot the same density curve for the CD3 marker using the normalised data.

```{r}
# Plot densities of CD3 for each image
fergusonSPE |> 
  join_features(features = rownames(fergusonSPE), shape = "wide", assay = "norm") |> 
  ggplot(aes(x = CD3, colour = imageID)) + 
  geom_density() + 
  theme(legend.position = "none")
```

Here, we can see that the normalised data appears more bimodal, and we can clearly observe a CD3+ peak at 1.00, and a CD3- peak at around 0.3. Image-level batch effects also appear to have been mitigated.

We can also visualise the effect of normalisation on the UMAP, which shows that the images now overlap with each other to a much greater extent.

```{r}
set.seed(51773)
# Perform dimension reduction using UMAP.
fergusonSPE <- scater::runUMAP(
  fergusonSPE,
  subset_row = ct_markers,
  exprs_values = "norm",
  name = "normUMAP"
)

someImages <- unique(fergusonSPE$imageID)[c(1, 5, 10, 20, 30, 40)]

# UMAP by imageID.
scater::plotReducedDim(
  fergusonSPE[, fergusonSPE$imageID %in% someImages],
  dimred = "normUMAP",
  colour_by = "imageID"
)
```

## scMerge: Combining multiple spatial datasets

A common question that pops up when analysing spatial datasets is:

Can I combine multiple spatial datasets?

## sessionInfo

```{r}
sessionInfo()
```
